---
title: "SOMMD Notebook"
output: html_notebook
---

This is A Notebook that explain how to use the SOMMD package to perform analysis of molecular dynamics simulations. 

Load SOMMD Package

```{r}
devtools::load_all("..")
```
Read simulation files from pdb and xtc with a for cycle:


```{r}
#Read the first replica
trj <- read.trj(trjfile=, topfile="../data/Medium_Dataset/ref.pdb")
#Append all other trj files
for(trj_file in list.files("../data/Medium_Dataset/", pattern = "*.xtc")[-1]){
  rep <- read.trj(trjfile=trj_file, topfile="../data/Medium_Dataset/ref.pdb")
  trj <- cat_trj(trj, rep)
}
```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
