---
title: "Clustering of MD trajectory"
output:
  pdf_document: default
  html_notebook: default
header-includes:
- \usepackage[table]{xcolor}
- \usepackage{multirow}
---

# 1.Beginning Cluster Analysis

## 1.1 Introduction

The **SOMMD** package in R offers utilities for the analysis of biomolecular simulations using **Self-Organizing Maps** (SOM). SOM is a type of artificial neural network widely used for detecting patterns in data. A key feature of SOM is its ability to reduce dimensionality by mapping high-dimensional data onto a grid, while maintaining topological relationships—i.e., similar data points are placed near each other on the map. This property makes SOM particularly useful for analyzing complex datasets such as molecular dynamics simulations.

The **SOMMD** package includes functionality for reading trajectory data in various formats, calculating descriptors, training SOMs, and analyzing pathways within molecular dynamics simulations.

The aim of this notebook is to introduce the process of training a SOM using the **SOMMD** package, in conjunction with the **kohonen** package. This guide will cover the steps necessary to train a SOM and interpret the results.

## 1.2 Getting Started

First, load the **SOMMD** package:

```{r}
library(SOMMD)
```

To explore all available functions within the SOMMD package, use the following command:

```{r}
help(package = "SOMMD")
```

For detailed help on any specific function, use:

```{r}
help(FunctionName)
```

## 1.3 Reading Example Trajectory Data

The SOMMD package comes with example datasets. For this tutorial, we will use a molecular dynamics trajectory of the FOXP1 DNA-binding domain. The trajectory is stored in the GROMACS XTC format, with solvent atoms excluded to reduce file size.

The following code downloads and unzips the dataset, setting the file paths for the trajectory (xtc) and reference (gro) structure files:

```{r}
utils::download.file("https://figshare.com/ndownloader/articles/26935432/versions/3", destfile = "FOXP1_dataset.zip", method = "curl")
utils::unzip("FOXP1_dataset.zip")

xtcfile <- "./01_FOXP1_dataset/FOXP1_WT-1-PROD.xtc"
strfile <- "./01_FOXP1_dataset/FOXP1_WT.gro"
```

Now, read the trajectory data into R using the read.trj function:

```{r}
trj <- read.trj(trjfile=xtcfile, topfile=strfile)
```

The [`read.trj`]{style="color:magenta"} function processes the input files and returns a new trj object. To view a summary of the object, you can use:

```{r}
print(trj)
```

This will display basic information about the system, including the number of atoms (1482 in this case) and the total number of frames in the trajectory (2001 frames). A key feature of the [`trj`]{style="color:magenta"} object is that it contains atomic information about the system, as read from the `topfile`{style="color:magenta"} argument (the reference structure file).

## 1.4 Training a SOM using atomic coordinates

Self-Organizing Maps can be trained using various descriptors. One of the most fundamental descriptors of a protein's conformation is its atomic coordinates. This descriptor will allow us to group similar conformations based on their structural similarity, typically measured by Root Mean Square Deviation (RMSD).

In this example, we will train the SOM using the C⍺ (alpha carbon) atomic coordinates. Since atomic coordinates can be sensitive to translational and rotational differences, we first align the frames to a reference structure to minimize such effects.

The following code selects the C⍺ atoms and performs the alignment:

```{r}
ca.inds <- which(trj$top$elety=="CA")
trj.fit <- fit.trj(trj, trj.inds=ca.inds)
```

Next, the atomic XYZ coordinates are extracted and passed to the SOM for training. The code below sets up a hexagonal grid and specifies the parameters for the SOM training process:

```{r}
SOM <- kohonen::som(trj2xyz(trj.fit, inds=ca.inds), 
                    grid = kohonen::somgrid(8, 8, "hexagonal", neighbourhood.fct="gaussian", toroidal=TRUE), 
                    dist.fcts="euclidean", rlen=500, mode='pbatch')
```

Here, we specify a toroidal map shape, which means that the map is periodic—i.e., neurons at the edges of the map are adjacent to those on the opposite side. This is particularly useful when analyzing data with cyclic or periodic properties.

To inspect the structure of the trained SOM, we can plot the U-Matrix, which visualizes the distance between neighboring neurons:

```{r}
plot(SOM, type = 'dist.neighbours', heatkey = TRUE, shape='straight', main="U-Matrix")
```

The U-Matrix plot displays the distance between neighboring neurons. Regions with high gradients (yellow/white areas) indicate a significant difference between adjacent neurons, whereas low gradients (red areas) represent neurons that are similar to each other. By examining this plot, you can gain insights into the clustering of the data and identify regions of the map that correspond to distinct conformational states.

## 1.5 SOM Training using distance matrix

An alternative and often more accurate descriptor of protein conformation is the **distance matrix**. Unlike atomic coordinates, which can be sensitive to the system's rotational and translational movements, the distance matrix is independent of these transformations. As a result, it does not require a frame superposition step, making it particularly useful for analyzing systems where rotational and translational motions are of concern.

To reduce the number of distances used as descriptors, only those between residues forming native contacts (i.e., distances between Cβ atoms < 1.1 nm) will be considered. Additionally, the analysis excludes the C- and N-terminal portions of the protein due to their inherent flexibility, which can introduce noise into the analysis.

The following code read the native structure as a gro file, compute the selection of distances forming a native contacts and then compute the descriptors that will be used to train the SOM.

```{r}
ref.str <- read.struct(strfile)
cb.inds <- which(trj$top$elety=="CB" & trj$top$resno > 470 & trj$top$resno < 542)
sele.dists <- native.cont(struct=ref.str, distance=1.1, atoms=cb.inds)
DIST <- calc.distances(trj, atoms=cb.inds, sele=sele.dists)
```

*The read.struct function reads the native structure from the gro file.
*The native.cont function analyzes the structure and returns an index of distances that are below the reference distance value (1.1 nm) in the native structure. Only these distances will be considered for the SOM training.
*The calc.distances function then computes the actual distances between the selected atoms (Cβ in this case) and stores them in the DIST object.

After computing the descriptors, the SOM is trained using these distance matrix values:

```{r}
SOM <- kohonen::som(DIST, grid = kohonen::somgrid(8, 8, "hexagonal", 
                                         neighbourhood.fct="gaussian", toroidal=TRUE), 
                    dist.fcts="euclidean", rlen=500, mode='pbatch')
```

As with the previous example, we can plot the U-Matrix to inspect the structure of the trained SOM:

```{r}
plot(SOM, type = 'dist.neighbours', heatkey = TRUE, shape='straight', main="U-Matrix")
```

The **U-Matrix** plot provides a visual representation of the SOM’s structure, similar to the previous example. This allows us to compare how the SOM trained with different features (atomic coordinates vs. distance matrix) has influenced the clustering of protein conformations.

To compare this new U-Matrix with the previous one, consider the following:

- **Regions with light colors** (extreme distance values) indicate that the selected features strongly differentiate certain conformations from others. These regions mark boundaries between clusters of conformations that are distinct in terms of the specific descriptor (e.g., distances between Cβ atoms).
- **Regions with darker colors** (low distance values) show areas where the conformations are similar according to the distance matrix. This suggests that those conformations are grouped together based on similar interactions or contact patterns between residues.

However, every study case is different, and you may encounter different patterns in the U-Matrix depending on your specific system and analysis. In some cases, you might see more defined clusters with extreme values, while in other cases, the U-Matrix may appear more "shaded," with less contrast between regions. This variability is expected and reflects how the SOM organizes the data based on the descriptors used.

## 1.6 Clustering of Neurons

In the context of a Self-Organizing Map (SOM), neurons are units that represent clusters of similar data points (in this case, molecular conformations or states). Each neuron on the map corresponds to a vector in the feature space (for example, atomic coordinates or distance matrices). After training the SOM, the neurons will form a grid where nearby neurons are more similar to each other than those farther apart.

To enhance the interpretation of the SOM, we can apply hierarchical agglomerative clustering to group neurons that are similar based on their feature vectors. Agglomerative clustering merges similar neurons into clusters. A crucial step in clustering is determining the optimal number of clusters. One way to do this is by looking at silhouette scores, which indicate how well-defined the clusters are.

### Visualizing and Choosing the Number of Clusters

The silhouette score measures how similar a neuron is to its own cluster compared to other clusters. Higher silhouette values indicate better-defined clusters. To visualize this, we first plot the silhouette scores for different numbers of clusters:

```{r}
par(mfrow=c(1,2))
#Plot the silhouette score
sil.score <- silhouette.score(SOM, clust_method="complete", interval=seq(2,30))
plot(sil.score, type='b', pch=19, lwd=1, xlab="Number of clusters", ylab='Average silhouettes')
for(i in seq(0, max(sil.score[,1]), by=2)){
    abline(v=i, col="grey", lwd=0.5)
}

#Plot the silhouette profile for 8 number of clusters
sil.prof <- silhouette.profile(SOM, Nclus=8, clust_method="complete")
plot(sil.prof, main="", do.clus.stat = FALSE, do.n.k = FALSE)
abline(v=mean(sil.prof[,3]), col='red', lty=3, lwd=2)
```

The first plot displays the average silhouette score for each number of clusters. The ideal number of clusters is typically the value that corresponds to the highest average silhouette score, but it’s also important to visually inspect the silhouette profiles to assess the quality of the clusters for specific numbers. This is shown in the second plot, where the silhouette profile for 8 clusters shows individual cluster quality. The red dashed line indicates the mean silhouette value for each cluster in this configuration. By inspecting both the silhouette score and the silhouette profiles, you can decide on the optimal number of clusters for your SOM.

### Dividing the SOM into Clusters

Once a reasonable number of clusters has been chosen, the SOM can be divided using the cutree function, which performs hierarchical clustering on the neurons. Below, we divide the SOM into 8 clusters and color the neurons according to their cluster assignments:

```{r}
#Divide the SOM in the selected number of clusters
Nclus <- 8
SOM.hc <- cutree(hclust(dist(SOM$codes[[1]], method="euclidean"), method="complete"), Nclus)
#Choose a pleasing to the eye set of colors
COL.SCALE <- c("#1f78b4", "#33a02c", "#e31a1c", "#ffff88", "#6a3d9a", 
               "#a0451f", "#96c3dc", "#fbb25c", "#ff7f00", "#bea0cc", 
               "#747474", "#f88587", "#a4db77")
#Plot the SOM colored by clusters
plot(SOM, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), shape='straight', main="")
kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=5)

```

In this code, the SOM is colored according to the cluster each neuron belongs to. The kohonen::add.cluster.boundaries function visually highlights the boundaries between clusters, making it easier to observe the groupings on the map.

### Visualizing Neurons and Clusters

After clustering, we can visualize the neurons and clusters with their corresponding numbers. This gives a clear identification of each neuron and cluster, enhancing interpretability:

```{r}
#Plot the SOM with neuron numbers
plot(SOM, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), shape='straight', main="")
kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=5)
som.add.numbers(SOM, scale=0.5, col="black")
som.add.clusters.legend(Nclus=8, color.scale=COL.SCALE)
```

Here, we use som.add.numbers to label each neuron with its corresponding number, and som.add.clusters.legend to display a legend indicating the cluster colors.

### Interpretation of Neurons and Clusters

* **Neurons**: Each neuron represents a specific feature vector (or a set of molecular properties), and its position on the map reflects its similarity to other neurons. Neurons can be seen as **microstates** of the system. Neurons close to each other on the map have similar characteristics, while neurons far apart differ significantly.
* **Clusters**: The clusters group neurons with similar characteristics together. The boundaries between clusters indicate transitions from one group of similar neurons to another. Clusters can be seen as **macrostates** of the system. The number of clusters you select can reveal different levels of organization in the data.

## 1.7 Extraction of representative frames

After clustering the neurons, it's useful to extract representative frames from the SOM to examine the specific conformations associated with each neuron or cluster. The following code extracts the representative frames for each neuron and cluster:

```{r}
#Get a vector of representative frames for each neuron
NEUR_repres <- neur.representatives(SOM)
cat("Frames representatives of neurons: ")
cat(NEUR_repres)
#Get representatives for each cluster
CL_repres <- cluster.representatives(SOM, SOM.hc)
cat("\n\nNeurons representatives of each cluster: ")
cat(CL_repres$neurons)
cat("\n\nFrames representatives of each cluster: ")
cat(CL_repres$frames)

```

If NA values appear in the output, they correspond to empty neurons or clusters that do not have a representative frame. This can occur if a cluster or neuron contains too few data points to produce a representative structure.
Once you've identified the representative frames, you can extract the corresponding conformation (structure) from the simulation trajectory. Use the `trj2pdb`{style="color:magenta"} function to save the structure of a representative frame in PDB format:

```{r}
#Estract the representative conformation of Neuron 8
trj2pdb(trj = trj, frame=NEUR_repres[8], filename = "./01_FOXP1_dataset/Neuron_8.pdb")

#Estract the representative conformation of Cluster B
trj2pdb(trj = trj, frame=CL_repres$frames["B"], filename = "./01_FOXP1_dataset/Cluster_B.pdb")
```

Note that `trj2pdb`{style="color:magenta"} saves a pdb file with the coordinate of a selected frame of the simulation. Here we are passing as input the frame of the representative structures.
The representative frames provide a way to visualize the most typical conformations for each neuron or cluster. By extracting these frames, you can compare how different groups of neurons (or clusters) correspond to specific structural configurations. This allows you to better understand the relationship between the SOM's clustering and the underlying protein dynamics.

## 1.8 Neurons Population

To visualize the population of each neuron, you can represent the number of frames associated with each neuron using circles. The size of each circle is proportional to the number of frames (or "population") that fall within that neuron. This can help you see which neurons are more populated, indicating that they are representative of a larger portion of the dataset.

```{r}
#Plot the SOM with circles with size proportional to the neuron population 
plot(SOM, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), shape='straight', main="")
kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=5)
population <- neur.population(SOM)
som.add.circles(SOM, population, scale=0.9)
```

Neurons with larger circles represent those that are populated by more frames. This can indicate that these neurons represent a more commonly occurring or stable conformation in your dataset. The `scale` argument in `SOM.add.circles`{style="color:magenta"} controls the size of the circles. You can adjust it to ensure that the circles are properly scaled and do not overlap or exceed the dimensions of the neurons.
  
## 1.9 Plot of System-Specific Properties

In addition to representing neuron population, you can visualize system-specific properties by using color scales to show the average value of a certain property for each neuron. This allows you to analyze how different properties vary across the map. For example, you can compute the RMSD (Root Mean Square Deviation) between the C- and N-terminal ends of a protein domain to represent the variability of protein conformations.

Here’s how you can read a file containing the RMSD values for each frame of the simulations, and visualize such a property:

```{r}

RMSD_file <- "./01_FOXP1_dataset/FOXP1_RMSD_WT-1-PROD.dat"
RMSD <- read.table(RMSD_file)
Neur.avg.RMSD <- average.neur.property(SOM, RMSD[,1])
par(mfrow=c(1,2))

#Plot the SOM with circles with size proportional to the average RMSD values
plot(SOM, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), shape='straight', main="")
kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=5)
som.add.circles(SOM, Neur.avg.RMSD, scale=0.5)
#Plot the SOM colored according to the average property value of each neuron
plot(SOM, type = "property", property=Neur.avg.RMSD, shape='straight', palette.name=colorRampPalette(c("blue", "green", "red")), main="")
kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=5)
```

In the first plot the SOM is shown with circles whose size is proportional to the average RMSD values of each neuron. Larger circles indicate neurons that correspond to conformations with higher RMSD, reflecting greater deviation.
In the second plot the SOM is colored according to the average RMSD values for each neuron. A color gradient from blue (low RMSD) to red (high RMSD) is used to show how the RMSD varies across the map.

You may notice that some neurons have very extreme values for RMSD. These extreme values could indicate conformations that are very different from the majority of the dataset. Such neurons might represent outliers or particular conformation sampled during MD simulations.

Always ensure that the length of the property vector matches the number of frames in your trajectory. If you are analyzing a different property (such as radius of gyration or solvent accessibility), the same principles apply: the length of the property vector should match the number of frames.

As an exercise, try to identify which motions or structural changes lead to the neuron or neurons with the highest RMSD values. What are the key differences in conformation between these neurons and others with lower RMSD? This can help you understand the dynamics and variability within the system.

# **2. Mapping data from multiple replicas**

When analyzing multiple simulations of the same system, it's useful to combine the trajectories into a single dataset to inspect potential differences and similarities across the simulations. This can be done using the `cat.trj`{style="color:magenta"} cat.trj function in the SOMMD package, which concatenates multiple simulations. In this example, we will compare five replicas of the same system.

First, we read all five xtc files corresponding to the replicas of the system. Make sure the trajectory files are correctly loaded with the associated gro file for the topology.

```{r}
#Read the xtc files in the folder
folder <- "./01_FOXP1_dataset"
xtcfiles <- list.files(folder, pattern = "FOXP1_WT-.-PROD.xtc")
strfile <- "./01_FOXP1_dataset/FOXP1_WT.gro"

MD1 <- read.trj(trjfile=paste(folder, "/", xtcfiles[1], sep=''), topfile=strfile)
MD2 <- read.trj(trjfile=paste(folder, "/", xtcfiles[2], sep=''), topfile=strfile)
MD3 <- read.trj(trjfile=paste(folder, "/", xtcfiles[3], sep=''), topfile=strfile)
MD4 <- read.trj(trjfile=paste(folder, "/", xtcfiles[4], sep=''), topfile=strfile)
MD5 <- read.trj(trjfile=paste(folder, "/", xtcfiles[5], sep=''), topfile=strfile)
```

Here, the five replicas are loaded one by one using the `read.trj`{style="color:magenta"} function, which takes in the trajectory file (xtc) and the topology file (gro).

Next, we concatenate the individual trajectories into one combined dataset using `cat.trj`{style="color:magenta"}.

```{r}
trj <- cat.trj(MD1, MD2, MD3, MD4, MD5)
```

The `cat.trj`{style="color:magenta"} function combines all the individual simulation trajectories into one large trajectory object. This will allow us to analyze all replicas together. 

**Note:** After concatenating, the `trj$start` and `trj$end` fields are updated, which will be useful later for mapping each frame to its original replica.

Once the concatenation is complete, remove the individual simulation objects to save memory. This step is optional and depend on the size of the input trajectory and on the RAM available on your machine.

```{r}
# Remove individual simulations to save memory
rm(MD1, MD2, MD3, MD4, MD5)
```

Now that we have a concatenated trajectory, we can proceed with training the SOM. In this case, we calculate the distances between specific atoms (for example, the C-beta atoms of a certain region of the protein) and use those distances as input for the SOM.

```{r}
ref.str <- read.struct(strfile)
cb.inds <- which(trj$top$elety=="CB" & trj$top$resno > 470 & trj$top$resno < 542)
sele.dists <- native.cont(struct=ref.str, distance=1.1, atoms=cb.inds)
DIST <- calc.distances(trj, atoms=cb.inds, sele=sele.dists)
SOM <- kohonen::som(DIST, grid = kohonen::somgrid(8, 8, "hexagonal", 
                                         neighbourhood.fct="gaussian", toroidal=TRUE), 
                    dist.fcts="euclidean", rlen=1000, mode='pbatch')
```

The distance matrix (DIST) is computed for a specific set of atoms. These distances are used to train the SOM, which clusters the frames of the concatenated trajectory into a map.

To determine the optimal number of clusters, we can calculate the silhouette score for different cluster numbers and plot the results.

```{r}
sil.score <- silhouette.score(SOM, clust_method="complete", interval=seq(2,30))
plot(sil.score, type='b', pch=19, lwd=1, xlab="Number of clusters", ylab='Average silhouettes')
for(i in seq(0, max(sil.score[,1]), by=2)){
    abline(v=i, col="grey", lwd=0.5)
}

```

Using hierarchical clustering, we cut the dendrogram into the chosen number of clusters (e.g., 8 clusters).

```{r}
NClus=8
SOM.hc <- cutree(hclust(dist(SOM$codes[[1]], method="euclidean"), method="complete"), Nclus)
```

Now that we have determined the clusters, we can visualize the SOM with the cluster information, adding neuron numbers and a cluster legend.

```{r}
COL.SCALE <- c("#1f78b4", "#33a02c", "#e31a1c", "#ffff88", "#6a3d9a", 
               "#a0451f", "#96c3dc", "#fbb25c", "#ff7f00", "#bea0cc", 
               "#747474", "#f88587", "#a4db77")

#Plot the SOM with neuron numbers
plot(SOM, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), shape='straight', main="")
kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=5)
som.add.numbers(SOM, scale=0.5, col="black")
som.add.clusters.legend(Nclus=8, color.scale=COL.SCALE)
```

This plot shows the SOM with each neuron numbered and colored according to its cluster assignment.
To inspect how each replica samples the SOM space, you can plot the population of each replica. This will show you which replicas are more representative of certain clusters and which ones sample unique regions.

```{r}
par(mfrow=c(2,3))

for(rep in 1:5){
  plot(SOM, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), shape='straight', main="")
  kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=3)
  population <- neur.population(SOM, start=trj$start, end=trj$end, N=rep)
  som.add.circles(SOM, population, scale=0.4)
}
```

This set of plots shows the population of each replica on the SOM map. The size of the circles represents how many frames from each replica fall into each neuron, helping to visualize the overlap and unique sampling of the space by different replicas.

**Excercise:** Select a cluster or neuron that is unique to a particular replica and investigate the structural differences that make it distinct from the others. This will allow you to understand how different replicas explore different conformations of the system.
