---
title: "Analisys of Pathways in protein unfolding simulations"
output:
  pdf_document: default
  html_notebook: default
header-includes:
- \usepackage[table]{xcolor}
- \usepackage{multirow}
---

# 1.SOM Training

## 1.1 Introduction

This notebook demonstrates how to use the SOMMD package to analyze pathways sampled during molecular dynamics (MD) simulations. Detailed instructions on how to perform SOM training and cluster analysis can be found in the Clustering of MD trajectory notebook.

Here, we focus on analyzing pathways sampled during Steered Molecular Dynamics (SMD) simulations of a protein domain unfolding process. SMD is an enhanced sampling technique that applies an external force to pull the system in a specific direction, allowing us to explore otherwise inaccessible conformational transitions. In this case, the pulling direction is defined by the distance between the C-terminal and N-terminal ends of the protein.

The system under investigation is the PAS-B domain of ARNT. For more details on the system and the interpretation of the results, refer to the following paper: [`S Motta et al., J. Chem. Theory Comput. 2021, 17, 4, 2080–2089 doi.org/10.1021/acs.jctc.0c01308`]{.underline}

## 1.2 Compute Descriptors

First, load the SOMMD package:

```{r} library(SOMMD)}
```

Next, download and extract the molecular dynamics (MD) trajectory files:

```{r}
#Dowload the MD simulation data 
utils::download.file("https://figshare.com/ndownloader/articles/26935465/versions/2", destfile = "ARNT_dataset.zip", method = "curl")
utils::unzip("ARNT_dataset.zip")
```

Now, load the PDB structure and trajectory files into R:

```{r}
#Read the names of the xtc files in the folder
strfile <- "./02_ARNT_dataset/ARNT.pdb"
folder <- "./02_ARNT_dataset"
xtcfiles <- list.files(folder, pattern = "*.xtc")

#Read the first replica
trj <- read.trj(trjfile=paste(folder, "/", xtcfiles[1], sep=''), topfile=strfile)
#Append all other trj files
for(trj_file in xtcfiles[-1]){
  rep <- read.trj(trjfile=paste(folder, "/", trj_file, sep=''),
                  topfile=strfile)
  trj <- cat.trj(trj, rep)
}
```

Note that the first trajectory file is loaded, and then a loop is used to sequentially load and concatenate all remaining trajectory files in the dataset.
This ensures that all simulation data is combined into a single dataset for further analysis.

To train a Self-Organizing Map (SOM), we must define descriptors that effectively capture the structural changes in the protein during unfolding. A useful choice is the pairwise distances between C𝛃 atoms, particularly those forming native contacts in the folded state. These distances provide insights into how the native structure is disrupted over time.
We can select these distances using the `native_contacts`{style="color:magenta"}function:

```{r}
#Read reference pdb with native conformation
ref.str <- read.struct("./02_ARNT_dataset/ARNT.pdb")
#Select only Cbeta atoms to perform the analysis
sele.atoms <- which(trj$top$elety=="CB")
#Choose only native contacts
sele.dists <- native.cont(struct=ref.str, distance=1.0, atoms=sele.atoms)
#Compute distances for SOM training. 
DIST <- calc.distances(trj, mol.2=FALSE, sele=sele.dists, atoms=sele.atoms)
```

**Why C𝛃 Pairwise Distances?**

* C𝛃 atoms are often used to describe protein conformational changes because they are centrally located in residues and less affected by fluctuations of hydrogen atoms.
* The selected distances correspond to native contacts, meaning they are particularly relevant for tracking the unfolding process.

Depending on the specific system and type of process you are interested in, you may want to choose a different set of descriptors.

## 1.3 SOM Training

With the selected descriptors, we can now train the Self-Organizing Map (SOM) using the `kohonen` package. Here we define a 8x8 grid of hexagonal neurons with a Gaussian neighborhood function.

```{r}
SOM <- kohonen::som(DIST, grid = kohonen::somgrid(8, 8, "hexagonal", 
                                         neighbourhood.fct="gaussian", toroidal=FALSE), 
                    dist.fcts="euclidean", rlen=1000, mode='pbatch')
```

To segment the SOM into meaningful groups of conformations, we perform hierarchical clustering and assess silhouette scores to determine the optimal number of clusters.

```{r}
sil.score <- silhouette.score(SOM, clust_method="complete", interval=seq(2,30))
plot(sil.score, type='b', pch=19, lwd=1, xlab="Number of clusters", ylab='Average silhouettes')
for(i in seq(0, max(sil.score[,1]), by=2)){
    abline(v=i, col="grey", lwd=0.5)
}
```

The **silhouette score** quantifies how well-separated each cluster is. The **optimal number of clusters** is identified as the peak in the silhouette score curve. Depending on the level of detail needed, one may stop at the first peak for a coarse clustering or consider subsequent peaks for a more refined partitioning of the data.

In this example, we assume an optimal number of 10 clusters, but since SOM training is stochastic, this value may change. Be sure to inspect your silhouette score plot and adjust accordingly.


```{r}
Nclus <- 10
#Divide the SOM in the selected number of clusters
SOM.hc <- cutree(hclust(dist(SOM$codes[[1]], method="euclidean"), method="complete"), Nclus)
```

Finally, we assign colors to the clusters and visualize the SOM map:

```{r}
#Choose a pleasing to the eye set of colors
COL.SCALE <- c("#1f78b4", "#33a02c", "#e31a1c", "#ffff88", "#6a3d9a", 
               "#a0451f", "#96c3dc", "#fbb25c", "#ff7f00", "#bea0cc", 
               "#747474", "#f88587", "#a4db77")
#Plot the SOM colored by clusters
plot(SOM, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), shape='straight', main="")
kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=5)
```

In this map each hexagon represents a neuron in the SOM, while the color coding corresponds to different conformational clusters. This visualization provides a structured representation of conformations sampled during the unfolding.

## 1.4 Representation of properties on SOM

To analyze how specific structural properties are distributed across the SOM, we will compute the distance between the C⍺ atoms of the N- and C-terminal ends of the protein domain and visualize this property on the map. This distance provides insight into whether the protein is in a more folded (shorter distance) or unfolded (longer distance) state.

```{r}
#Select the index of the first and last CA atoms
Terminals <- c(head(which(trj$top$elety=="CA"),1), tail(which(trj$top$elety=="CA"),1))
#Compute distances between these two atoms in every frame of the simulation
Term_dist <- apply(trj$coord[Terminals,,], 3, dist)
#Compute average property value for each neuron
Neur.avg.d <- average.neur.property(SOM, Term_dist)

#Plot the SOM colored according to the average property value of each neuron
plot(SOM, type = "property", property=Neur.avg.d, shape='straight', palette.name=colorRampPalette(c("blue", "yellow", "red")), main="")
kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=5)
```

In the resulting SOM, neurons are colored according to the average C⍺-C⍺ terminal distance of the structures they represent. The blue regions correspond to low distance values, indicating that the protein is in a more compact (folded) conformation. Conversely, the red regions represent high distance values, suggesting that the protein is in a more extended (unfolded) state. Often the two extremes are positioned at opposite corners of the map.

By inspecting the spatial organization of these states on the SOM, and comparing with the clusters, you can gain insights into the structural transitions explored during the simulation and how different conformational states are grouped.

# 2.Analysis of Pathways

## 2.1 Trace Pathways

Each trajectory represents a sequence of transitions between neurons, revealing dominant unfolding/folding or binding pathways. Identifying the most frequent pathways can provide insights into conformational changes and preferred energy landscapes.
To analyze the pathways explored by different replicas on the SOMwe can use the `trace_path`{style="color:magenta"} function, which visualizes the trajectory followed by a simulation on the SOM.

To simplify the plotting of pathways from different replicas, the `trj` object stores the start and end frames (`trj$start` and `trj$end`) of each replica—merged using the
`cat_trj`{style="color:magenta"} function. Using this information, we can plot the pathway of a specific replica as follows:

```{r}
#Plot the SOM colored by clusters
par(mfrow=c(2,2))
for(rep in c(1,3,9,11)){
  plot(SOM, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), 
       shape='straight', main=paste("Replica ", rep, sep=""))
  kohonen::add.cluster.boundaries(SOM, SOM.hc, lwd=3)
  #Add the path tracing
  trace.path(SOM, start=trj$start, end=trj$end, N=rep, pts.scale=0.5, lwd.scale=0.4)
}
```

This allows us to visualize the movement of each replica across the SOM, helping to identify similarities and differences in trajectories.

## 2.2 Clustering of Pathways

To better understand the types of pathways sampled, we can perform pathway clustering using:

```{r}
path.clust <- cluster.pathways(SOM, start=trj$start, end=trj$end, time.dep="dependent")
plot(path.clust, xlab="")

```

This analysis groups similar pathways across different replicas. Here, we used the option: `time.dep="dependent"`, , which compares frames at the same time points in different simulations. This approach is valid for Steered MD simulations with a constant pulling speed, as they progress at the same rate.

However, if the simulations evolve at different speeds (e.g., in constant-force Steered MD), the `time.dep="independent"` setting should be used instead, as it allows for time-independent pathway comparisons.

## 2.3 Mapping new data

A trained SOM can also be used to analyze new simulation datasets. This is done using the function `map.data`{style="color:magenta"}, which assigns new data points to the closest neurons in the existing SOM.

In this example, we analyze data from additional Steered MD simulations run at lower pulling speeds, stored in a different folder. We first load these simulations and compute the same descriptors used to train the original SOM.

```{r}
#Dowload the MD simulations at lower pulling speed 
utils::download.file("https://figshare.com/ndownloader/articles/26935495/versions/2", destfile = "ARNT_lowspeed_dataset", method = "curl")
utils::unzip("ARNT_lowspeed_dataset.zip")

#Read the names of the xtc files in the folder
strfile <- "./02_ARNT_lowspeed_dataset/ARNT.pdb"
folder <- "./02_ARNT_lowspeed_dataset"
xtcfiles <- list.files(folder, pattern = "*.xtc")

#Read the first replica
trj2 <- read.trj(trjfile=paste(folder, "/", xtcfiles[1], sep=''), topfile=strfile)
#Append all other trj files
for(trj_file in xtcfiles[-1]){
  rep <- read.trj(trjfile=paste(folder, "/", trj_file, sep=''),
                  topfile=strfile)
  trj2 <- cat.trj(trj2, rep)
}

#Compute distances (the same used for SOM training)
DIST2 <- calc.distances(trj2, mol.2=FALSE, sele=sele.dists, atoms=sele.atoms)
```

Now, we map this new dataset onto the previously trained SOM:

```{r}
#Map new data on the existing SOM
SOM_new <- remap.data(SOM=SOM, X=DIST2)
```

The returned `SOM_new` oobject contains the original SOM structure, now enriched with the newly mapped data.

To check whether the new data is well represented by the original SOM, we compare the distribution of distances between data points and their assigned SOM neurons. Low values indicate that frames are well represented, while higher values suggest poor representation.

```{r}
#Map new data on the existing SOM
boxplot(SOM$distances, xlim=c(0.5,2.5), ylim=c(0,50), pch=16, ylab="Distance")
boxplot(SOM_new$distances, at=2, add=TRUE, pch=16)
axis(1, at=c(1,2), labels=c("High Speed", "Low Speed"))
```

From this plot, we observe that the distance distributions are similar, meaning that the new data is consistent with the original dataset. A slight increase in distances is expected, as the SOM was initially trained only on the high-speed dataset.

Since the mapped SOM contains both datasets, we can now visualize the pathways sampled at lower pulling speeds:

```{r}
par(mfrow=c(2,3))
for(rep in c(1:5)){
  plot(SOM_new, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), 
       shape='straight', main=paste("Low Speed Replica ", rep, sep=""))
  kohonen::add.cluster.boundaries(SOM_new, SOM.hc, lwd=3)
  trace.path(SOM_new, start=trj2$start, end=trj2$end, N=rep, pts.scale=0.5, lwd.scale=0.5)
}
```

From these plots, we can compare the pathways explored at different pulling speeds. As can be seen the pathways sampled by the low-speed replicas are similar to those sampled by higher-speed simulations. This indicates that, despite the different pulling speeds, the overall exploration of the conformational space remains similar, suggesting that the system may converge toward similar structural states under varying conditions.

