---
title: "Transition network analysis in ligand-protein metadynamic simulation"
output:
  pdf_document: default
  html_notebook: default
header-includes:
- \usepackage[table]{xcolor}
- \usepackage{multirow}
---

# 1. SOM Training

## 1.1 Introduction

This notebook demonstrates how to use the SOMMD package to analyze pathways sampled during molecular dynamics simulations and construct a network graph representing transitions between neurons. Detailed instructions on SOM training and path tracing can be found in the previous notebooks. Here, we focus on building a network to represent pathways explored during a metadynamics simulation of a protein-ligand binding event.

Metadynamics is an enhanced sampling MD method in which a Gaussian-shaped potential is periodically added to bias the system at the current position of predefined collective variables (CVs). This technique allows the system to escape local minima and explore new regions in the CV space. In this case, the CVs used for metadynamics are the position along a predefined path S(x) and the deviation from the reference path Z(x).

The system under investigation is a 1.8 µs metadynamics simulation of the THS-020 ligand binding to the PAS-B domain of HIF-2⍺. For further details on the system and the interpretation of the analysis, refer to the following papers:

`[L.` `Callea et al., J. Chem. Theory Comput. 2021, 17, 7, 3841–3851 doi.org/10.1021/acs.jctc.1c00114]`, `[S Motta et al., J. Chem. Theory Comput. 2022, 18, 3, 1957–1968 doi.org/10.1021/acs.jctc.1c01163]`.

## 1.2 Train SOM with native interactions

We begin by loading the SOMMD package and reading the simulation files (GRO and trajectory files).

```{r}
library(SOMMD)
#Dowload the Metadynamic simulation and other files to execute the Notebook 
utils::download.file("https://figshare.com/ndownloader/articles/26935510/versions/1", destfile = "HIF_dataset.zip", method = "curl")
utils::unzip("HIF_dataset.zip")

#Read the xtc file 
xtcfile <- "./03_HIF_dataset/HIF_MetaD.xtc"
strfile <- "./03_HIF_dataset/HIF.gro"
trj <- read.trj(trjfile=xtcfile, topfile=strfile)
```

To analyze the ligand binding process, we need meaningful descriptors that capture the ligand’s position within the binding site. A practical approach is to consider intermolecular distances between ligand and protein atoms. Here, we select the heavy atoms forming a contact (distance \< 6 Å) in the reference structure.

```{r}
#Read reference pdb with native conformation
ref.str <- read.struct(strfile)
#Select protein and ligand atoms
protein.sele <- which(ref.str$atom$resid!="020")
ligand.sele <- which(ref.str$atom$resid=="020")
#Select heavy atoms
heavy.atoms <- which(base::startsWith(ref.str$atom$elety, "H")==FALSE)
#Choose only native contacts
sele.dists <- native.cont(struct=ref.str, distance=0.6, mol.2=ligand.sele, atoms=heavy.atoms)
#Compute distances for SOM training. 
DIST <- calc.distances(trj, mol.2=ligand.sele, sele=sele.dists, atoms=heavy.atoms)

print(dim(DIST))
```

The computed distance matrix consists of 629 descriptors across 9175 simulation frames. This matrix serves as input for training the SOM using the kohonen package. We define a 10×10 non-periodic SOM with hexagonal neurons.

```{r}
SOM_nat <- kohonen::som(DIST, grid = kohonen::somgrid(10, 10, "hexagonal", 
                                         neighbourhood.fct="gaussian", toroidal=FALSE), 
                    dist.fcts="euclidean", rlen=500, mode='pbatch')
```

To determine the optimal number of clusters, we evaluate the silhouette score for different cluster numbers and select the one maximizing the average silhouette value.

```{r}
sil.score <- silhouette.score(SOM_nat, clust_method="complete", interval=seq(2,30))
plot(sil.score, type='b', pch=19, lwd=1, xlab="Number of clusters", ylab='Average silhouettes')
for(i in seq(0, max(sil.score[,1]), by=2)){
    abline(v=i, col="grey", lwd=0.5)
}

```

Assuming 10 as a optimal number of clusters according to the silhouette profile (see previuos notebooks), we assign colors to the clusters and visualize the SOM map:

```{r}
Nclus <- 10
#Divide the SOM in the selected number of clusters
SOM.hc <- cutree(hclust(dist(SOM_nat$codes[[1]], method="euclidean"), method="complete"), 10)
#Choose a pleasing to the eye set of colors
COL.SCALE <- c("#1f78b4", "#33a02c", "#e31a1c", "#ffff88", "#6a3d9a", 
               "#a0451f", "#96c3dc", "#fbb25c", "#ff7f00", "#bea0cc", 
               "#747474", "#f88587", "#a4db77")
#Plot the SOM colored by clusters
plot(SOM_nat, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), shape='straight', main="")
kohonen::add.cluster.boundaries(SOM_nat, SOM.hc, lwd=5)
```

Finally, we analyze how the CVs S(x) and Z(x) distribute over the trained SOM, allowing us to interpret the relationship between sampled metastable states and the simulation’s biased variables.

```{r}
#Plot the value of S(x) and Z(x) over the SOM.
Sx_file <- "./03_HIF_dataset/HIF_MetaD_Sx.dat"
Sx <- read.table(Sx_file)
Neur.avg.Sx <- average.neur.property(SOM_nat, Sx[,1])

Zx_file <- "./03_HIF_dataset/HIF_MetaD_Zx.dat"
Zx <- read.table(Zx_file)
Neur.avg.Zx <- average.neur.property(SOM_nat, Zx[,1])

par(mfrow=c(1,2))
#Plot the SOM colored according to the average property value of each neuron
plot(SOM_nat, type = "property", property=Neur.avg.Sx, shape='straight', palette.name=colorRampPalette(c("blue", "green", "red")), main="S(x)")
kohonen::add.cluster.boundaries(SOM_nat, SOM.hc, lwd=5)
#Plot the SOM colored according to the average property value of each neuron
plot(SOM_nat, type = "property", property=Neur.avg.Zx, shape='straight', palette.name=colorRampPalette(c("blue", "green", "red")), main="Z(x)")
kohonen::add.cluster.boundaries(SOM_nat, SOM.hc, lwd=5)
```

These plots allow us to interpret the SOM by mapping neurons and clusters onto the Free Energy Surface (FES) obtained from metadynamics, helping to identify different states and their relationships within the binding process.

## 1.3 Train SOM using a selection of atoms and a distance cap

Now, let’s adopt a different approach by selecting atoms that are particularly relevant to the binding process. These atoms should describe both the binding site and the entrance of the binding pocket. Ideally, the selection should include backbone atoms as well as side chains that are large, polar, or charged—especially when side-chain dynamics and interactions play a key role in binding. Likewise, the ligand atom selection should capture the core molecular structure along with important functional groups.

Here, we use the same atom set from our reference study:
`[S Motta et al., J. Chem. Theory Comput. 2022, 18, 3, 1957–1968 doi.org/10.1021/acs.jctc.1c01163]`. Additionally, we apply a capping value of 1.2 in the `calc.distances`{style="color:magenta"} function, which ensures that all distances greater than 12 Å are set to 12 Å. This helps reduce noise by making all unbound states appear similar, allowing the SOM to focus on the conformations where the ligand is in contact with the protein.

```{r}
#atoms selection
sele.atoms <- c(127, 157, 167, 214, 223, 253, 263, 624, 668, 688, 699, 793, 798, 807, 816, 838, 852, 865, 871, 880, 905, 910, 919, 924, 1011, 1025, 1046, 1052, 1059, 1076, 1087, 1121, 1130, 1315, 1323, 1569, 1603, 1608, 1630, 1635, 1781, 1792, 1793, 1794, 1795)

#Select ligand atoms
ligand.sele <- which(ref.str$atom$resid=="020")

#Compute distances for SOM training. 
DIST <- calc.distances(trj, mol.2=ligand.sele, atoms=sele.atoms, cap=1.2)

print(dim(DIST))
```

The selection `sele.atoms` contains the indexes of protein and ligand atoms deemed important for our analysis. In this case, we have 40 protein atoms and 5 ligand atoms. Then, the `calc.distances`{style="color:magenta"} function computes all intermolecular distances between every protein and ligand atom pair, yielding 40 × 5 = 200 distance values. These computed distances will now serve as input for training the SOM.

```{r}
SOM_sel <- kohonen::som(DIST, grid = kohonen::somgrid(10, 10, "hexagonal", 
                                         neighbourhood.fct="gaussian", toroidal=FALSE), 
                    dist.fcts="euclidean", rlen=500, mode='pbatch')
```

To determine the optimal number of clusters, we compute the silhouette score for different cluster numbers.

```{r}
sil.score <- silhouette.score(SOM_sel, clust_method="complete", interval=seq(2,30))
plot(sil.score, type='b', pch=19, lwd=1, xlab="Number of clusters", ylab='Average silhouettes')
for(i in seq(0, max(sil.score[,1]), by=2)){
    abline(v=i, col="grey", lwd=0.5)
}

```

Assuming 12 as a optimal number of clusters according to the silhouette profile (see previuos notebooks), we assign colors to the clusters and visualize the SOM map:

```{r}
Nclus <- 12
#Divide the SOM in the selected number of clusters
SOM.hc <- cutree(hclust(dist(SOM_sel$codes[[1]], method="euclidean"), method="complete"), Nclus)
#Choose a pleasing to the eye set of colors
COL.SCALE <- c("#1f78b4", "#33a02c", "#e31a1c", "#ffff88", "#6a3d9a", 
               "#a0451f", "#96c3dc", "#fbb25c", "#ff7f00", "#bea0cc", 
               "#747474", "#f88587", "#a4db77")
#Plot the SOM colored by clusters
plot(SOM_sel, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), shape='straight', main="")
kohonen::add.cluster.boundaries(SOM_sel, SOM.hc, lwd=5)

```

### Mapping Free Energy Surface Properties to the SOM

Next, we map the free energy landscape obtained from metadynamics onto the SOM, using the S(x) and Z(x) properties.

```{r}
#Re-compute the average per-neuron value of Sx and Zx for the new SOM
Neur.avg.Sx <- average.neur.property(SOM_sel, Sx[,1])
Neur.avg.Zx <- average.neur.property(SOM_sel, Zx[,1])

par(mfrow=c(1,2))
#Plot the SOM colored according to the average property value of each neuron
plot(SOM_sel, type = "property", property=Neur.avg.Sx, shape='straight', palette.name=colorRampPalette(c("blue", "green", "red")), main="S(x)")
kohonen::add.cluster.boundaries(SOM_sel, SOM.hc, lwd=5)
#Plot the SOM colored according to the average property value of each neuron
plot(SOM_sel, type = "property", property=Neur.avg.Zx, shape='straight', palette.name=colorRampPalette(c("blue", "green", "red")), main="Z(x)")
kohonen::add.cluster.boundaries(SOM_sel, SOM.hc, lwd=5)
```

**Key Insights from This SOM Approach**: 
By using this refined atom selection and distance capping, we observe two major advantages:

* **Improved Handling of the Unbound State**: The region of the SOM corresponding to the completely unbound state (S(x) > 8) is now limited to a few neurons in one corner of the map. This is the results from the imposed capping value, which prevents excessive focus on trivial unbound conformations.
* **Better Description of Bound Conformations**: Bound states (S(x) < 4) now occupy an entire side of the SOM rather than being confined to a corner. This suggests that we have improved resolution in describing different binding poses.

## 1.4 Trace Pathways

At this stage, we can visualize the sampled binding/unbinding pathways over the SOM.

```{r}
#Plot the SOM colored by clusters
plot(SOM_sel, type = "mapping", bgcol=COL.SCALE[SOM.hc], col=rgb(0,0,0,0), 
       shape='straight', main="")
kohonen::add.cluster.boundaries(SOM_sel, SOM.hc, lwd=3)
trace.path(SOM_sel, pts.scale=1, lwd.scale=0.5)

```

However, the resulting plot is quite complex because the simulation exhibits frequent transitions back and forth along the reaction coordinate.

## **1.5 Representing SOM as a graph network**

By utilizing the transition matrix derived from the simulation data, it is possible to construct a graph network that represents all the possible pathways sampled during the simulations. This approach provides a visual and quantitative method to explore the dynamics between different states of the system.

The first step is to compute the transition matrix, which encodes the probabilities of transitioning between different states (or nodes) in the SOM. The function `comp.trans.mat`{style="color:magenta"} computes this matrix based on the SOM structure.

```{r}
#Compute transition matrix
Tr_mat <- comp.trans.mat(SOM_sel, start = 1)
#Create igraph object
net <- matrix2graph(Tr_mat, SOM_sel, SOM.hc, COL.SCALE, diag=FALSE)
```

In the `comp.trans.mat`{style="color:magenta"} function, the `start` argument specifies the starting frame of the simulation. If you have multiple simulation replicas, you should account for the transition between the last frame of one replica and the first frame of the next. This is done using the trj\$start vector that contains the starting frames of all the replicas concatenated with the `cat.trj`{style="color:magenta"} function. In this case, we are using a single simulation, so the starting point is simply the first frame (1).

Once the transition matrix is created, you can visualize the network. The `igraph` package is used to create and plot the network.

```{r}
edge.start <- igraph::ends(net, es=igraph::E(net), names=F)[,1]
edge.col <- igraph::V(net)$color[edge.start]
#Plot network with the SOM layout
plot(net, edge.arrow.size=igraph::E(net)$width/5, edge.curved=0.17, edge.color='black', vertex.label="", layout=SOM_sel$grid$pts)
```

This code visualizes the network using the default SOM layout, which places the nodes based on their positions in the SOM grid. This layout is straightforward, but you may prefer to use other algorithms that can optimize the layout of the network based on the edges.

To better visualize the relationships between the states, you can apply a force-directed layout algorithm. The `Fruchterman-Reingold` algorithm is commonly used for this purpose. It optimizes the placement of nodes to reduce edge crossings and improve the readability of the graph.

```{r}
coords = igraph::layout_with_fr(net)
plot(net, edge.arrow.size=igraph::E(net)$width/5, edge.curved=0.17, edge.color='black', vertex.label="", layout=coords)

```

Note that the `layout_with_fr()`{style="color:magenta"} function is not deterministic, meaning that each time you run the algorithm, the node positions may change. You can try running the function multiple times to find a clear and stable layout.

### Community Detection: Kinetic-like Clustering

Another interesting approach is to perform community detection on the graph. This method identifies groups of nodes (communities) that are more densely connected to each other than to the rest of the network. In the context of SOM, this can be interpreted as finding clusters of states that are more likely to transition between each other.

The `walktrap` algorithm in `igraph` is useful for identifying communities in bidirectional graphs. It works by simulating random walks on the graph and detecting areas where the walks are more likely to stay within the same community.

```{r}
CL_walktrap <- igraph::cluster_walktrap(net)
COL <- COL.SCALE[igraph::membership(CL_walktrap)]
par(mfrow=c(1,2))
plot(SOM_sel, type = "mapping", bgcol=COL, col=rgb(0,0,0,0), shape='straight', main="")
kohonen::add.cluster.boundaries(SOM_sel, SOM.hc, lwd=3)
plot(net, edge.arrow.size=igraph::E(net)$width/10, edge.curved=0.17, vertex.color=COL, edge.color='black', vertex.label="", layout=coords)

```

Here, the network nodes (representing the states) are grouped based on the probability of transition between them, rather than their geometric similarity. The coloring of nodes reflects their community affiliation. You should easily identify two pathways that connect the unbound state with the two different bound states.

### Mapping Properties onto the Graph

You can also map specific properties of the states (e.g., S(x), Z(x)) onto the graph nodes to visualize how these properties vary across different states. This is done by assigning a color scale based on the property values.

For example, to visualize how the property S(x) varies across the graph, you can use the following code:

```{r}
#Set the color scale
col.palette <- colorRampPalette(c("blue", "green", "red"))(200)
#Convert the average property to the colors

par(mfrow=c(1,2))
#Do the plot for Sx
COL <- map.color(Neur.avg.Sx, col.palette)
plot(net, edge.arrow.size=igraph::E(net)$width/10, edge.curved=0.17, vertex.color=COL, edge.color='black', vertex.label="", layout=coords)

#Do the plot for Zx
COL <- map.color(Neur.avg.Zx, col.palette)
plot(net, edge.arrow.size=igraph::E(net)$width/10, edge.curved=0.17, vertex.color=COL, edge.color='black', vertex.label="", layout=coords)
```

In these plots, you will be able to clearly distinguish two distinct pathways: one leading to the native bound state (characterized by low S(x) and Z(x) values), and another leading to an alternative bound state (with low S(x) and high Z(x) values).

The network representation offers a powerful advantage when visualizing complex pathways. During simulations, you can sample various pathways, and the graph provides a unique visualization that integrates all these pathways into a single, comprehensive representation. This makes it easier to understand the transitions and relationships between states, and to identify key pathways that might not be immediately obvious from the raw simulation data.
